Fast gravitational wave parameter estimation without compromises

### 2023-07-21 src/tex/ms.tex
Add discussion related EOB

\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{pyRed}{RGB}{214, 39, 40}
\newcommand{\te}[1]{\textbf{\color{pyRed}(TE: #1)}}
\newcommand{\mi}[1]{\textsf{\color{teal}[\textbf{MI:} #1]}}

@@

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\software{LALInference} \citep{Veitch:2014wba}, \software{PyCBC Inference}
\citep{Biwer:2018osg}, and \software{Bilby}
\citep{Ashton:2018jfp,Romero-Shaw:2020owr}.  These packages have been tested by a
number of groups and are well regarded as standard tools. However, while these
tools have passed many robustness tests, they are known to be computationally
intensive. The exact amount of time needed to analyze one event depends on
factors like the duration and frequency of the signal, as well as features of
the specific waveform model. Typical runtimes for production-level analyses can
range from hours to weeks.

@@

To obtain a reference waveform, we currently use the \software{differential
evolution} algorithm \citep{Storn1997DifferentialE} available in the
\software{scipy} package \citep{2020SciPy-NMeth} to find the waveform parameters
which maximize the likelihood.  The reference waveform could also be produced
from trigger parameters precomputed by a search pipeline without additional
computation. Once we have obtained a reference waveform, we can check the accuracy
of the heterodyned likelihood by comparing its value to the original likelihood
at several points in the parameter space. We can then choose the number of bins such
that the difference between the values of the two likelihoods is smaller than
chosen tolerance threshold.

@@

% More sample helps training normalizing flow
Yet, in our work, the use of accelerators provides two independent advantages
that tremendously benefit the parameter estimation process. First, using
accelerators allows us to run many independent MCMC chains simultaneously, which
benefits the training of the normalizing flow. Since we generate the data we use
to train the normalizing flow on the fly, the more independent data we can feed
to the training process, the higher chance the normalizing flow can learn a
reasonable representation of the target distribution.  If we only used a small
number of chains, we would be limited to the correlated samples from each chain
and we would have to run more sequential steps to obtain the same amount of
independent samples in the end---with more chains the problem becomes
parallelizable and we can obtain the same number of training samples sooner. In
other words, being able to use many independent chains helps the normalizing
flow approximating the target density faster in wall time.

@@

\begin{figure*}
    \includegraphics[width=0.99\textwidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (blue) and \software{Bilby} (gray).
    }
    \label{fig:GW150914}
    \script{plot_GW150914.py}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.99\textwidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (blue) and \software{Bilby} (gray).
}
\label{fig:GW170817}
    \script{plot_GW170817.py}
\end{figure*}

For comparison, we produce equivalent runs with \software{Bilby}, using the same
exact data and priors.  We use the \software{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github \url{https://github.com/kazewong/TurboPE/tree/main/src}.  We
carry out these runs using \software{parallel Bilby} (\software{pBilby})
\citep{Smith:2019ucc} to distribute the computation over 400 Intel Skylake CPUs
for each event.  For the specific settings chosen, the wall-time duration of
each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day for GW170817.

@@

Figs.~\ref{fig:GW150914} and \ref{fig:GW170817} show that our posteriors are
consistent with those produced by \software{Bilby}.  For a quantitative
comparison, we compute the Jensen-Shannon divergence between our code and
\software{Bilby} for the marginalized distribution for each parameter. The
Jensen-Shannon divergence (JSD) is a symmetric measure of the distance between
two probability distributions, with a value of 0 indicating identical
distributions and a value of $\ln{2}$ nat representing the maximum
possible divergence between two distributions. The JSD values for the two events
are shown in Table \ref{tab:JSD}. The maximum JSDs for GW150914 and GW170817 are
\jsdMaxBBH and \jsdMaxBNS, and the mean JSDs are \jsdAvgBBH and \jsdAvgBNS,
respectively. The JSD values are comparable to those reported in
\cite{Romero-Shaw:2020owr}, which show our code agrees with existing tools.

\begin{table}[hbt!]
    \begin{center}
    \variable{output/js_table.tex}
    \caption{JSD values for the marginalized distribution of each parameter for
    GW150914 and GW170817 between our code and \software{Bilby}. The bold values
    indicate the parameters with the largest JSD.}
    \label{tab:JSD}
    \end{center}
\end{table}

@@

The two avenues discussed here (machine learning and reparameterizations)
represent two orthogonal directions one can take in building next generation PE
tools. On the one hand, there are modern techniques such as deep learning that
are very flexible and powerful, but rely on having highly robust training data.
On the other hand, there are traditional tools that make use of our
understanding of the underlying physics to ease sampling, which relies
on having good intuition of the problem and tends to not generalize.
Our method achieves the key advantages of both approaches, without many of their limitations.
%% [MI: the following is too vague, and we always rely on a reasonable prior,
%% so I'm removing it.]
% Both approaches rely on having a somewhat reasonable prior to approach the problem.

@@

% Additional traits for having differentiable waveforms
We are currently working on a number of improvements and extensions to our
current infrastructure. While the \textsc{IMRPhenomD} waveform approximant is a
reasonable start, it lacks some qualitative features that other
state-of-the-art models have, such as precession, subdominants moments of the
radiation, tides, or eccentricity. It also has a higher mismatch with reference
numerical relativity waveforms compared to more recent waveform models.
Currently, we are working on building differentiable implementations of
\textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb}, a precessing successor to
\textsc{IMRPhenomD} including subdominant harmonics, as well as the numerical
relativity surrogate waveforms, including \textsc{NRSur7dq4}
\citep{Varma:2018mmi}. Going forward, we expect the use of autodifferentiation
environments like \software{Jax} to become more prevalent in the waveform
development community, increasing the number of differentiable waveform models
available. This would not only be beneficial for parameter estimation, but also
for a number of other applications such as Fisher matrix computations, template
placement and calibrating waveforms to numerical relativity \citep{Coogan:2022qxs,Iacovelli:2022bbs,Iacovelli:2022mbg,ripplepaper}.

@@

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\software{differential evolution} algorithm we currently use has not been
implemented in \software{Jax}, and the \software{Jax} waveform we use is not
compatible with the parallelization scheme in the \software{scipy} library,
maximizing the likelihood currently takes us around 1 minute for GW170817. There
are two ways to reduce this time.

>>>>

\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\definecolor{pyRed}{RGB}{214, 39, 40}
\newcommand{\te}[1]{\textbf{\color{pyRed}(TE: #1)}}
\newcommand{\mi}[1]{\textsf{\color{teal}[\textbf{MI:} #1]}}
\newcommand{\corr}[1]{\textbf{\color{pyRed}(Correction: #1)}}

@@

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\software{LALInference} \citep{Veitch:2014wba}, \software{PyCBC Inference}
\citep{Biwer:2018osg}, and \textsc{Bilby}
\citep{Ashton:2018jfp,Romero-Shaw:2020owr}.  These packages have been tested by a
number of groups and are well regarded as standard tools. However, while these
tools have passed many robustness tests, they are known to be computationally
intensive. The exact amount of time needed to analyze one event depends on
factors like the duration and frequency of the signal, as well as features of
the specific waveform model. Typical runtimes for production-level analyses can
range from hours to weeks.

@@

To obtain a reference waveform, we currently use the \textsc{differential
evolution} algorithm \citep{Storn1997DifferentialE} available in the
\software{scipy} package \citep{2020SciPy-NMeth} to find the waveform parameters
which maximize the likelihood.  The reference waveform could also be produced
from trigger parameters precomputed by a search pipeline without additional
computation. Once we have obtained a reference waveform, we can check the accuracy
of the heterodyned likelihood by comparing its value to the original likelihood
at several points in the parameter space. We can then choose the number of bins such
that the difference between the values of the two likelihoods is smaller than
chosen tolerance threshold.

@@

% More sample helps training normalizing flow
Yet, in our work, the use of accelerators provides two independent advantages
that tremendously benefit the parameter estimation process. First, using
accelerators allows us to run many independent MCMC chains simultaneously, which
benefits the training of the normalizing flow. Since we generate the data we use
to train the normalizing flow on the fly, the more independent data we can feed
to the training process, the higher chance the normalizing flow can learn a
reasonable representation of the target distribution.  If we only used a small
number of chains, we would be limited to the correlated samples from each chain
and we would have to run more sequential steps to obtain the same amount of
independent samples in the end---with more chains the problem becomes
parallelizable and we can obtain the same number of training samples sooner. In
other words, \corr{being able to use many independent chains helps the normalizing
flow approximating the target density faster in wall time.}

@@

\begin{figure*}
    \includegraphics[width=0.99\textwidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (blue) and \textsc{Bilby} (gray).
    }
    \label{fig:GW150914}
    \script{plot_GW150914.py}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.99\textwidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (blue) and \textsc{Bilby} (gray).
}
\label{fig:GW170817}
    \script{plot_GW170817.py}
\end{figure*}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \software{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github \url{https://github.com/kazewong/TurboPE/tree/main/src}.  We
carry out these runs using \textsc{parallel Bilby} (\software{pBilby})
\citep{Smith:2019ucc} to distribute the computation over 400 Intel Skylake CPUs
for each event.  For the specific settings chosen, the wall-time duration of
each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day for GW170817.

@@

Figs.~\ref{fig:GW150914} and \ref{fig:GW170817} show that our posteriors are
consistent with those produced by \textsc{Bilby}.  For a quantitative
comparison, we compute the Jensen-Shannon divergence between our code and
\textsc{Bilby} for the marginalized distribution for each parameter. The
Jensen-Shannon divergence (JSD) is a symmetric measure of the distance between
two probability distributions, with a value of 0 indicating identical
distributions and a value of $\ln{2}$ nat representing the maximum
possible divergence between two distributions. The JSD values for the two events
are shown in Table \ref{tab:JSD}. The maximum JSDs for GW150914 and GW170817 are
\jsdMaxBBH and \jsdMaxBNS, and the mean JSDs are \jsdAvgBBH and \jsdAvgBNS,
respectively. The JSD values are comparable to those reported in
\cite{Romero-Shaw:2020owr}, which show our code agrees with existing tools.

\begin{table}[hbt!]
    \begin{center}
    \variable{output/js_table.tex}
    \caption{JSD values for the marginalized distribution of each parameter for
    GW150914 and GW170817 between our code and \textsc{Bilby}. The bold values
    indicate the parameters with the largest JSD.}
    \label{tab:JSD}
    \end{center}
\end{table}

@@

The two avenues discussed here (machine learning and reparameterizations)
represent two orthogonal directions one can take in building next generation PE
tools. On the one hand, there are modern techniques such as deep learning that
are very flexible and powerful, but rely on having highly robust training data.
On the other hand, there are traditional tools that make use of our
understanding of the underlying physics to ease sampling, which relies
on having good intuition of the problem and tends to not generalize.
Our method achieves the key advantages of both approaches, without many of their limitations.

@@

% Additional traits for having differentiable waveforms
We are currently working on a number of improvements and extensions to our
current infrastructure. While the \textsc{IMRPhenomD} waveform approximant is a
reasonable start, it lacks some qualitative features that other state-of-the-art
models have, such as precession, subdominants moments of the radiation, tides,
or eccentricity. It also has a higher mismatch with reference numerical
relativity waveforms compared to more recent waveform models. Currently, we are
working on building differentiable implementations of \textsc{IMRPhenomXPHM}
\citep{Pratten:2020ceb}, a precessing successor to \textsc{IMRPhenomD} including
subdominant harmonics, as well as the numerical relativity surrogate waveforms,
including \textsc{NRSur7dq4} \citep{Varma:2018mmi}. Going forward, we expect the
use of autodifferentiation environments like \software{Jax} to become more
prevalent in the waveform development community, increasing the number of
differentiable waveform models available. This would not only be beneficial for
parameter estimation, but also for a number of other applications such as Fisher
matrix computations, template placement and calibrating waveforms to numerical
relativity
\citep{Coogan:2022qxs,Iacovelli:2022bbs,Iacovelli:2022mbg,ripplepaper}.
\corr{Another family of waveform models that could benefit from this approach is
the effective-one-body (EOB) models \citep{Buonanno:1998gg,PhysRevD.89.061502}.
However, the EOB family requires solving time-dependent ODEs to compute the
waveform and the truncation of the ODEs is dynamically determined at runtime.
JIT in \software{Jax} requires the computational graph of a function to be fixed
at compile time, which means the ODEs solving part of EOB is not compatible with
\software{Jax}, so the models in the EOB family may not benefit as much as the
other families.}

@@

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\textsc{differential evolution} algorithm we currently use has not been
implemented in \software{Jax}, and the \software{Jax} waveform we use is not
compatible with the parallelization scheme in the \software{scipy} library,
maximizing the likelihood currently takes us around 1 minute for GW170817. There
are two ways to reduce this time.
